---
title: "blog_script"
author: Joseph Mack
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Background
This post, written by Joseph Mack, explores the working of the unified
interpolation/extrapolation tools described in Colwell et al. 2012 *Journal of
Plant Ecology* "Models and estimators linking individual-based and sample-based
rarefaction, extrapolation and comparison of assemblages" doi:
10.1093/jpe/rtr044. Joseph shared this tutorial with the Fall 2021 Biodiversity
Measurement seminar.

# Joseph's tutorial

First, we must import a sample presence-absence dataset.

```{r}
spp_incidence <- read.csv("data/ann_incidence.csv", row.names = 1)
```
The equations used in this analysis are derived from a Bernoulli Product model.
Under this model, species $i$ has probability theta of occurring. What is the
probability that species $i$ is not present in a given site?

Do we know this probability for each species in our dataset?

To proceed, we must determine how many species (`Sobs`) and sites (`sites`) are
in the data set (`Sobs`)?

```{r Explore dimensions}
sites <- length(spp_incidence)
Sobs <- nrow(spp_incidence)
sites
Sobs
```
The equation below can be found on pg. 6 of Colwell et al. 2012. It calculates
the incidence-based frequency of a given species.

```{r incidences}

#What is the incidence-based frequency (Yi) of Sp1?
Ysp1 = sum(as.numeric(spp_incidence["sp1",]))
Ysp1

#What about the other species? 
spp_names = rownames(spp_incidence)
Yi <- c(1:12)
for (i in 1:12) {
  Yi[i] <- sum(spp_incidence[i,])
}

```

The code below creates a table of incidence frequencies (the row sums of our
data set), which will be used to calculate `Qk`.

```{r cbind data}
incidence_freqs <- data.frame(cbind(Yi, spp_names))

```

Next, we must determine how many 'unique' species are in the dataset. In other
words, how many species occur in only one site? This requires us to calculate
`Qk`, the incidence frequency count and an essential statistic for sample-based
extrapolation.

```{r Q(k) returns number of k-tons}
Q <- function(k) {
  out = 0
  for (i in 1:length(incidence_freqs$Yi)) {
    if (incidence_freqs$Yi[i] == k) {
      out = out + 1
    } 
  }
  return(out)
}

#How many species are found in only 1 site? 4 sites? 9 of the sites? All the sites?
Q(1)
Q(4)
Q(9)
Q(18)

```

Below is a function to interpolate/rarefy the sample dataset (equation 17 in
Colewll et al.). It computes the number of species we would recover if we
sampled from `x` fewer sites (`x < sites`).

```{r Rarefy sites}
Ssample_rare <- function(x) {
  for (i in 1:Sobs) {
    out <- (choose((sites - as.numeric(incidence_freqs$Yi[i])), x))/
      choose(sites, x)
  }
  return(Sobs - sum(out))
}

#How many species would we recover if we sampled from only 1 site? 5 sites? 9 sites?
Ssample_rare(1)
Ssample_rare(5)
Ssample_rare(9)
```

The values for `Ssample_rare` don't look right. This rarefaction function is
broken! Can you figure out why? Hint: what is `[i]` doing in the code below?

The rarefaction code below works! Why?

```{r Rarefy sites fixed}
Ssample_rare <- function(x) {
  out = c()
  for (i in 1:Sobs) {
  out[i] <- (choose((sites - as.numeric(incidence_freqs$Yi[i])), x))/choose(sites, x)
}
  return(Sobs - sum(out))
}

#NOW, how many species would we recover if we sampled from only 1 site? 5 sites? 9 sites?
Ssample_rare(1)
Ssample_rare(5)
Ssample_rare(9)
#These values make more sense!
```

Now that we have a means to interpolate our data, we can make a rarefaction plot.

```{r first rarefaction plot}
diversity_r = sapply(1:18, Ssample_rare) # This outputs a vector of Ssample_rare 
# outputs for 1 to 18 sampling sites. 
plot(1:18
     , diversity_r
     , type = "l"
     , ylab = "Number of Species"
     , xlab = "Number of Sites")
```

Next, we can work on extrapolating the dataset `(Ssample(sites + s))` using
equations 18 and 21 from Colwell et al. 2012. This addresses how many species we
would find if we sampled `s` additional sites.

To do this, we first need to estimate `Q(0)`: the number of species absent from
our samples.

```{r Chao2 estimation}
#Because Q(2) > 0 for our data set, we can use the following equation based on the Chao2 estimator:
Q0 = ((sites-1)/sites)*((Q(1)^2)/(2*Q(2)))
Q0
```

With a value for `Q0`, we can implement the authors' equation for extrapolation
as a function of `s` additional samples:

```{r extrapolation function}
Ssample_extr <- function(s) {
  out <- Sobs + (Q0*(1 - (1 - (Q(1)/(Q(1) + (sites*Q0))))^s))
  return(out) 
}

#How many species would we recover if we sampled 2 extra sites? 50 extra sites?
Ssample_extr(2)
Ssample_extr(50)

#Make an extrapolation plot for 72 additional sites. 
diversity_e = sapply(1:72, Ssample_extr) #This outputs a vector of Ssample_extr 
# outputs for 1 to 50 additional sampling sites. 
plot(1:72
     , diversity_e
     , type = "l"
     , lty = 2
     , ylab = "Number of species"
     , xlab = "Number of Additional Sites")
```

Now, we can combine the rarefaction and extrapolation plots into a single plot:

```{r SAC plot code}
library(ggplot2)
combined_diversity = c(diversity_r, diversity_e) #Combining the vectors of 
# rarefied and extrapolated species counts, for 1 - 90 sites.
combined_diversity
sampling_sites = 1:90 #vector for 18 rarefied sites + 72 additional extrapolated 
# sites
sampling_sites
diversity_est <-
  data.frame(cbind(combined_diversity, sampling_sites))
sac_plot = ggplot(
  data = subset(diversity_est, sampling_sites <= 18),
  aes(x = sampling_sites, y = combined_diversity)
) + 
  geom_line(size = 0.75) +
  geom_line(
    data = subset(diversity_est, sampling_sites >= 18),
    aes(x = sampling_sites, y = combined_diversity),
    lty = 2,
    size = 0.75
  ) +
  geom_point(aes(x = 18, y = 12), size = 3, pch = 16) +
  xlab("Number of Sampling Sites") +
  ylab("Number of Species") +
  ggtitle("Species Accumulation Curve") +
  theme(plot.title = element_text(size = 8.5))
```

```{R print SAC plot}
sac_plot
```

What does the point represent in this plot? The dotted line? The solid line? Is
our data set a good approximation for the total number of species in the region
sampled?

It is possible to obtain a 95% confidence interval about the species
accumulation curve. However, the variance estimators provided by Colwell et al.
2012 are difficult to understand. As an alternative, we can employ
non-parametric bootstrapping to get 95% confidence intervals about the curve.
The code below is not explicitly described in Colwell et al. 2012, but it
employs the same estimators we used above.

Below is a function for getting bootstrap replicates of our rarefaction and
extrapolation estimates.

```{r rarefaction bootstrapping}
Ssample_boot <- function(x, s) {
  # First, we need to build matrices and vectors to hold our boot-strapped data replicates. 
  boot_data <- matrix(nrow = 12, ncol = 18) 
  Yi_boot <- c(1:12)
  for (i in 1:12) {
    # The line below creates a replicate dataset from our original spp_incidence dataset, 
    # with replacement. 
    boot_data[i, ] <- rbind(as.numeric(sample(spp_incidence[i, ]
                                             , size = 18
                                             , replace = TRUE))) 
    #The for loop below calculates the incidence frequencies from the replicated dataset. 
    for (j in 1:12) {
      Yi_boot[j] <- sum(boot_data[j,])  
    }
  }
  # Below is the equation for rarefaction that we implemented earlier, where x is 
  # a sample smaller than sites. 
  out_rare = c()
  for (i in 1:Sobs) {
    out_rare[i] <- choose(sites - Yi_boot[i], x)/choose(sites, x)
  } 
  
  #This calculates a value of Q2 for the replicate data set. 
  Q2 = sum(Yi_boot == 2) 
  
  #This calculates a value of Q1 for replicate data set. 
  Q1 = sum(Yi_boot == 1) 
  
  # We are using a different equation for Q0 here (Equation 22), 
  # to avoid dividing by 0 when Q2 is 0 in some of our replicates. 
  Q0_b = ((sites - 1)/sites) * (Q1 * ((Q1 - 1))/(2 * (Q2 + 1))) 
  
  # Below is just equation 18, with some special conditions to account for 
  # scenarios where the function divides by 0. 
  out_extr <- Q0_b * (1 - (exp((-s * Q1))/(Q1 + sites * (Q0_b)))) 
  if (is.nan(out_extr) == TRUE) {
    return(c(Sobs - sum(out_rare), Sobs))
  } else {
  return(c(Sobs - sum(out_rare), Sobs + out_extr))
  } 
}
```

The code below tests that the function is working. The first variable (`x`)
specifies the sample size to be interpolated/rarefied, while the second variable
(`s`) specifies the additional sites to be extrapolated. The output should be
different each time you run it.

```{r test boot}
Ssample_boot(1,2) 
```

The code below gets bootstrap replicates using `replicate` and `mapply`. How
many bootstrap replicates are we running?

```{r set up bootstrap}
var1 <- 1:18
var2 <- 1:90
boot_data <- replicate(100, mapply(Ssample_boot, var1, var2))
```

Now that we have our bootstrap replicates, we can set up an empty matrix to
store our bootstrap replicate data. Each column is a site (1 - 18 interpolated,
19 - 90 extrapolated), and each row is a bootstrapped replicate.

```{r store boot data}
combined_boot_data <- matrix(ncol = 90, nrow = 100) 

#The code below populates the matrix. 
for (i in 1:100) {
  combined_boot_data[i, 1:18] <- boot_data[, , i][1, 1:18]
}
for (i in 1:100) {
  combined_boot_data[i, 19:90] <- boot_data[, , i][2, 19:90]
}
combined_boot_data <- data.frame(combined_boot_data)
```

Now that we have bootstrapped replicates for rarefaction and extrapolation, we
can estimate mean, 2.5% , and 97.5% quantiles for each site (1 - 18 rarefied, 19
- 90 extrapolated). These values can then be used to estimate a 95% confidence
interval centered about our sample data.

```{r summarys stats}
#This creates an empty vector to store the means of our bootstrapped data. 
boot_means <- c() 

#This estimates the average across 100 bootstrap replicates for each site (1 - 90).
for (i in 1:90) {
  boot_means[i] = (sum(combined_boot_data[,i]))/100
} 

#This creates an empty data frame to store the 2.5% and 97.5% quantiles of our bootstrapped data. 
boot_quantiles <- matrix(nrow = 2, ncol = 90)
boot_quantiles <- data.frame(boot_quantiles) 

# This computes the 2.5% and 97.5% quantiles of our data set, 
# populating the data frame we created above. 
for (i in 1:90) {
  boot_quantiles[,i] <- unname(quantile(combined_boot_data[,i], probs = c(0.025, 0.975)))
}
CI_data <- rbind(combined_boot_data, boot_means, boot_quantiles)
```

With the average, 2.5%, and 97.5% quantiles, we can compute the lower and upper
confidence bounds for each rarefied and extrapolated site.

For each site, the lower confidence bound = sample data - (bootstrap average -
2.5% bootstrap quantile).

```{r lower bound}
LCB <- c()
for (i in 1:90) {
    LCB[i] <- combined_diversity[i] - (CI_data[101,i] - CI_data[102,i])
}
```

For each site, the upper confidence bound = sample data + (97.5% bootstrap
quantile - bootstrap average).

```{r upper bound}
UCB <- c()
for (i in 1:90) {
    UCB[i] <- combined_diversity[i] + (CI_data[103,i] - CI_data[101,i])
}
confidence_intervals <- data.frame(cbind(LCB,UCB))
```

With this, we can plot 95% confidence intervals about the species accumulation
curve using the LCB and UCB values calculated above. The block of code below is
mostly identical to the plotting section above. What did we change?

```{r plot SAC with bounds}
sampling_sites = 1:90 #18 sites + 72 additional sites
diversity_est <-
  data.frame(cbind(combined_diversity, sampling_sites))
sac_plot = ggplot(
  data = subset(diversity_est, sampling_sites <= 18),
  aes(x = sampling_sites, y = combined_diversity)
) +
  geom_ribbon(
    confidence_intervals,
    mapping = aes(ymin = LCB, ymax = UCB),
    fill = "skyblue",
    col = "blue"
  ) + #Use geom_ribbon to plot our upper and lower confidence intervals.
  geom_line(size = 0.75) +
  geom_line(
    data = subset(diversity_est, sampling_sites >= 18),
    aes(x = sampling_sites, y = combined_diversity),
    lty = 2,
    size = 0.75
  ) +
  geom_point(aes(x = 18, y = 12), size = 3, pch = 16) +
  xlab("Number of Sampling Sites") +
  ylab("Number of Species") +
  ggtitle("Species Accumulation Curve") +
  theme(plot.title = element_text(size = 8.5))
sac_plot
```